\section{Conclusions and Discussion}
\label{sec:conclusion}

At this point we return to the question posed at the beginning of this paper:
\textbf{Are smartphone channel scans useful?} We believe that the answer is
\textbf{yes.} Through three extensive case studies, we have demonstrate that
smartphone measurements can contribute unique insights about large scale
wireless network deployments which are difficult or impossible to glean from
site surveys, statically deployed sniffers or infrastructure side logs. Quite
simply, smartphones represent real network users, and their data is
representative in the way that these other sources of measurements cannot be.

\begin{comment}
To conclude, we address several practical issues in collecting smartphone
scan data: (1) Will smartphone scan data compromise user privacy? (2) What
are the incentives for user to contribute data? and (3) Does device
heterogeneity affect the usefulness of the measurements? 

\begin{table*}[t!]
  \centering
  \begin{tabular}{cccccccc}
    \input{./figures/privacy.tex}
  \end{tabular}
  \caption{\textbf{Summary of Data Required for Each Case Study.} $\times$: Not
  required. %$\Diamond$: Optional. $\surd$: Required.}
}
  \label{tab:privacy}
  \vspace*{\aftercaptiongap}
\end{table*}

\subsection{Privacy}
\label{subsec:privacy}

One particular concern that arises when sharing channel scan data is privacy.
To address this, we first point out the types of user data that the
infrastructure side can already collect. Then we discuss the minimum
requirements in terms of channel scan data for each case study. The
discussion also serves as guidelines to our data anonymization process for
releasing the UB-Scan and ND-Scan datasets.

In enterprise network environments, since all user devices have to be
authenticated to associate to the APs, the \wifi{} session information,
including user's identity, device MAC address and association time, is
recorded by the management software. Therefore, a complete view of the user's
\wifi{} connection activity is already available on the infrastructure side.
Even when the user's device is not connected to any AP, the infrastructure
may still be able to detect the presence of the device by sniffing probe
packets\footnote{iOS 8 introduces a MAC randomization feature in probe
packets to preserve user's location privacy.}. Against this backdrop,
providing channel scan information does not represent an additional loss of
privacy compared to what infrastructure networks can already monitor about
their users.

Next we discuss the minimum requirements in terms of scan data for each of the
case studies in the previous sections. These requirements are summarized in
Table~\ref{tab:privacy}.

\begin{enumerate}
    %
  \item For all case studies, device identification information is not
    utilized in any way; thus, it can be stripped off even before scan
    results data are uploaded.
%
\item Timestamp information is not needed either, as long as the multiple
  scan result entries can be correctly grouped together and be identified as
  from one single scan. In practice, however, coarse grained time information
  such as year and month can be kept for post processing convenience.
%
\item Since the channel scan data are mainly used to help monitor central
  managed networks with predetermined SSIDs, such information is not required
  from the scan results. Furthermore, the user can choose only to upload
  channel scan data when they contain a predefined set of SSIDs, to avoid
  revealing information such as home AP SSID. 
%
\item Signal strength information is necessary in analyzing and predicting
  the device's association behavior in the spatial planning case study, yet
  is optional in the other two studies. Additionally, this information can be
  replaced by an ordering of the APs by signal strength, eliminating the need
  for absolute RSSI values.
%
\item Channel information is not required, since the infrastructure has the
  complete knowledge of the channel assignments for all APs at any instant.
%
\item \wifi{} connection information is only useful in identifying the
  association choice from the scan results in the spatial planning case
  study. If such information can be annotated in the scan results before data
  uploading, \wifi{} connection information is not required either. For
  example, the device can identify the scans happened just before a \wifi{}
  connection event, and mark the corresponding BSSID as chosen AP in the scan
  result.
%
\end{enumerate}

To summarize, only the BSSID information is necessary for the infrastructure to
identify the APs. Other potentially sensitive information can either removed
or anonymized.

\subsection{Incentives}
\label{subsec:incentive}

We next discuss incentives for user to participate in data collection. First, as
we showed earlier in Section~\ref{sec:scan}, smartphones already perform \wifi{}
scans aggressively, thus the natural scan rate already provides a stream of
network measurements with sufficient temporal granularity for monitoring
purposes. By only passively harvesting the scan results, the overhead incurred
in the process is kept low. And further steps can be taken to reduce the energy
consumption (e.g., only upload data when charging) and possibly metered cellular
data usage (e.g., only upload through \wifi{} networks).  Additionally, user
privacy can be maintained using the anonymization techniques we discussed in
Section~\ref{subsec:privacy}.

Then we look into incentives that encourage users to share the data. Note that
in enterprise networks, such data collection consent can be incorporated as part
of the authentication process. For instance, users may be required to install
the data collection app when connecting to the enterprise network, and the
installation can then allow the user's other devices to connect to the network.
For public \wifi{} service providers such as Boingo~\cite{boingo}, extra Quality
of Service (QoS) or price discounts can be offered to incentivize
participation.

In summary, the measurement overhead can be significantly reduced by passive
data collection and asynchronous delay tolerant data uploading, and effective
incentive mechanisms can encourage measurement participation.

\subsection{Device Heterogeneity}
\label{subsec:heterogeneity}

In the \ubscan{} dataset that is mostly analyzed throughout this paper, all scan
results are collected from devices of the same model (Nexus 5). We now reflect
on whether this hardware homogeneity is necessary, and what the impact of device
heterogeneity would be on each of the case studies.

\wifi{} chipsets and drivers from various manufacturers may have different
policies on the scanning rate. Yet, as we showed earlier in
Section~\ref{sec:scan}, most wireless devices from different vendors perform
\wifi{} scans quite aggressively. Therefore, the device heterogeneity shall
have negligible impact on the temporal resolution of the measurements.

On other hand, \wifi{} chipsets do have various RF characteristics, such as
carrier sensing range or antenna sensitivity. For instance, when building
client-perceived conflict graph (\S\ref{subsec:channel}), a more sensitive
device may reveal more conflict edges than a less sensitive one. And
similarly, devices with different sensitivity levels may disagree on the
offending set of the same RAP (\S\ref{subsec:rogue}).

To tolerate such diversity, the measurement post-processing logic can either
be aggressive by adopting the superset of observations from different
devices, or be conservative and only consider common conflicts reported by
most devices. Additionally, note that relative ordering of the APs by
signal strength is usually consistent for each device, therefore the device
heterogeneity has minimal impact on the spatial planning case study, where
only the relative ordering of the APs instead of the absolute RSSI values is
utilized.

\subsection{Measurement App}

Inspired by our results, we are developing a lightweight scan collection app
and set of analysis tools that we will make available on the Google Play
store and to network operators interested in using scans to better
understand their enterprise \wifi{} networks.
\end{comment}
