\section{Introduction}
\label{sec-introduction}

As mobile wireless devices proliferate and the demands they place on wireless
networks continue to grow, monitoring the health and performance of large
scale wireless networks is both essential and increasingly challenging. While
site surveys and infrastructure-side data collection both play important
roles in network monitoring, neither of these approaches can fully and
continuously reflect actual network conditions experienced by the clients the
network is intended to serve---particularly the mobile devices that generate
most wireless traffic. Site surveys are snapshots that are neither spatially nor
temporally representative, and AP-side data collection can only
measure wireless conditions at AP fixed locations, not at the changing
locations of mobile clients. As a result, multiple previous research studies
have demonstrated the value of client-side measurements to improve network
performance~\cite{mishra2005weighted,mishra2006client}, and the 802.11k
amendment~\cite{80211k} provides a mechanism allowing APs to collect this
information from clients to improve radio resource management.

Given the established interest in and emerging support for client-side wireless
network measurement, it is surprising that today we are discarding a
potentially-valuable source of client-side information: the large number of
channel scans performed continuously by mobile devices such as smartphones.
These devices naturally perform frequent channel scans to cope with
rapidly-fluctuating network environments caused by mobility. For example,
Android smartphones scan every 15~s when unassociated~\cite{hanover2014cutting}
to search for APs to connect to, and continue to scan every 60~s even when
associated to identify better APs as conditions change due to client mobility,
network interference, or other factors.

Considered as network measurements, channel scans share all the typical
benefits of client-side measurements---such as capturing the real conditions
experienced by end users, providing data that is impossible to gather using
APs alone, and being simpler and more representative than site surveys. In
addition, they have several other unique benefits. First, note that clients
perform these channel scans in order to effectively use wireless networks, not
to measure them. As a result, the measurements themselves are free and the
client overhead required to use them for monitoring purposes is limited to
the costs of temporary storage (which is minimal) and telemetry (which can be
minimized through delay tolerant and energy neutral data collection). Second,
in many cases channel scans remain useful even after being stripped of
information such as timestamps or device identifiers that could threaten
client privacy and potentially limit the willingness of users to share this
information.

All \wifi{} clients perform channel scans, but smartphones have several
advantages that make them uniquely suited to collecting and providing
measurements for enterprise networking monitoring. First, unlike laptops,
smartphones are always on and so scan continuously both during and between
periods of interactive use, providing better temporal resolution than devices
that are regularly powered down. Second, being highly portable, users are
likely to carry smartphones with them most of the time, giving these devices
the ability to observe more of the network than would be seen by stationary
or less-portable devices and making their measurements more representative of
all locations where mobile users might utilize the wireless network. Finally,
smartphone platforms already provide interfaces allowing apps (with
appropriate permissions) to access channel scans, and app marketplaces
provide an easy way to deploy the simple monitoring software required to
recover them to billions of mobile devices. Compared to other mobile devices,
smartphones make it (1) easy to collect (2) an enormous number of channel
scans from (3) everywhere that users might use multiple wireless networks.

\textbf{But whether (and how) this data can be made useful?} Can it contribute valuable
insights that would be impossible to gain from existing sophisticated
infrastructure-based approaches? That is the question that we set out to
answer in this paper. Our goal is to determine whether the billions of
discarded smartphone channel scans represent a missed opportunity or
redundant information that we can continue to safely ignore.

\sloppypar{%
  To do so, we take a walk on the client side. We utilize two large
  channel scan datasets collected by two smartphone testbeds (\PhoneLab{}:
  \num{5373682}~scans, 254~devices, 5~months; and \NetSense{}:
  \num{32564809}~scans, 125~devices, 32~months), described in more detail in
  Section~\ref{sec:dataset}. To examine the case for client-side measurement, we
  ignore analyses that could be performed using data from APs, including APs
  that utilize extra radio hardware to allow them to continuously perform
  AP-side measurements without disrupting normal client traffic. We also focus
  our studies on aspects of network performance and behavior of interest to
  network operators, not just scientists, since we anticipate that large-scale
  collection of channel scans will only take place if the data provides meaningful
  operational insights.
}

Our paper presents two case studies in Section~\ref{sec:case} using one or
both of our channel scan datasets:

\begin{itemize}
  \item \textbf{Section~\ref{subsec:channel}} looks at how the channel conflict
    graph differs from the client perspective, helping operators assess the
    effectiveness of infrastructure-only channel assignment algorithms at
    utilizing available spectrum.

  \item \textbf{Section~\ref{subsec:spatial}} investigates how load can be
    shifted among neighbor APs when a particular AP is not available, both for
    load balancing purpose and to evaluate network redundancy, allowing network
    operators to make better spatial planning decisions.
\end{itemize}

Section~\ref{sec:related} discusses related work before we present and
discuss our conclusions in Section~\ref{sec:conclusion}. We believe that our
case studies demonstrate the value of smartphone channel scans, and will
also publish the datasets so that others can extend our analyses or reevaluate
our conclusion.
